Leadership across cultures and contexts: an exploratory analysis

### 2020-05-12 leadership_across_cultures_contexts.Rmd
Edits in conclusion

Overall, our results suggest that although high status leaders are common across cultures and all levels of social organization, only some leaders are high status, namely, those with a limited set of qualities such as wealth, experience, decision-making abilities, intelligence, and inter-personal skills, as well as older age and conformity to social norms.

@@

**Dominance and prestige**. Much of the evolutionary theoretical literature on leadership has suggested social influence is most often achieved via one of two distinct strategies: dominance or prestige [for review see @garfield_review_2019; @garfield_evolutionary_2019]. There is a long history of leadership styles being described as based on either force and strength (i.e., dominance) or respect and special skills (i.e., prestige). Such dual-models are rooted in the coupling of ethology and ethnography [see @mead_sex_1935; @sahlins_poor_1963; @lewis_leaders_1974; @tiger_dominance_1970; @bernstein_dominance_1976]. Synthesizing the ethnographic and ethological evidence, @kracke_force_1978 linked "force and persuassion" to psychological-affect between leaders and followers in small-scale societies and @barkow_prestige_1980 suggested the evolution of prestige was linked to sexual selection and symbolic culture. The dominance-prestige theory as formulated by @henrich_evolution_2001 significantly advanced evolutionary approaches to leadership and social hierarchy and inspired much subsequent work across the behavioral sciences [e.g., @van_vugt_dual_2019; @cheng_two_2013; @maner_2016; @halevy_status_2012; @suessenbach_dominance_2019; @maner_dominance_2017; @garfield_evolutionary_2019; @garfield_investigating_2019]. @van_vugt_dual_2019 [p. 1], for example, "argue that there are two leadership styles: prestige style and dominance style," and @cheng_two_2013 [p. 119] in focusing on distinctions between dominance and prestige suggests, "human social stratification is characterized by the co-occurrence of both strategies." Exploratory cluster analyses reported here, however, suggest multiple features of leader qualities and leader functions, beyond dominance and prestige.

<!-- DISCUSSES COMPETENCIES FEATURE -->
**Leader capacities**. The individual-level competencies feature was comprised of variables associated with dominance-based leadership, including coercive authority and aggressiveness [@tiger_imperial_1971; @barkow_prestige_1975; @barkow_darwin_1989; @henrich_evolution_2001; @cheng_two_2013] as well as the qualities of knowledgeable or intelligent and experienced or accomplished, traits that have been emphasized by the prestige theory [@henrich_evolution_2001; @cheng_two_2013] and the more general computational services theory [@hagen_leadership_2019; @garfield_evolutionary_2019], and which are among the most strongly supported in the ethnographic record (Figure \@ref(fig:varsupport)). Interestingly, being feared and having supernatural qualities clustered more closely with the prestige-style traits than with the dominance-style traits. There is likely a strong connection between shamanism and ritual-based influence with both prestige and dominance styles of leadership. Ritual and supernatural capacities may represent a dimension of leader qualities in which dominance and prestige are coupled forming an alternative style of leadership, as shamans often provide valued public good, have special expertise and abilities, are feared, and achieve or maintain influence via threats or use of direct and/or indirect aggression. Artistic performance is also implicated in this style of leadership and ceremonial, artistic performances by shamans are common displays which often incorporate the superhuman abilities of shamans [@singh_cultural_2017].

As brief examples of the diversity of leader capacities, of the very many in the ethnographic record, among the Azande, "though the most knowledgeable witch-doctors tend to become leaders, there is no institutionalized organization or leadership among them" [@fo07-056, p. 85]; and among the Highland Scots, @es10-003 [p. 147] notes, "on the whole, their long ties with place and people, coupled with education, made them useful leaders of a community in which leaders were needed to adapt life to a changing economy." Such exceptionally talented leaders, however, also face unique challenges, particularly within more egalitarian societies which emphasizes personal autonomy and resit hierarchy in the context of cultural transitions. @nm09-056 [p. 183] describes Laura Cornelius Kellogg (also known as Minnie and Wynnogene, 1880-1947), a land-claims activist, writer, and perhaps the most famous female leader among the Iroquois:

@@

<!-- DISCUSSES PROSOCIALITY FEATURE -->
**Prosocial norms**. The prosociality feature of leader qualities, including generosity, interpersonal and oratory skills, and charisma as well as the role of cultural norms in shaping leadership styles, including proper behavior and having a favorable personality, generally correspond to a charismatic style of leadership that has been identified in industrialized settings and widely discussed in sociology and more recently emphasized in evolutionary psychology models of leadership [@den_hartog_culture_1999; @weber_distribution_1978; @grabo_charismatic_2016; @howell_role_2005]. The importance of prosociality  also corresponds to the @henrich_big_2015 Big-man mechanism, which suggests positive assortment of followers to prosocial leaders can facilitate the evolution of cooperative social norms and prestige-based leadership.
<!-- A few variables did not seem to fit prosocial theme, however, such as *Physical health*, *Physical formidable*, and *Drug consumption* (although drug use is often an important social activity). -->

@@

<!-- DISCUSSES BOTH MANAGEMENT FEATURES -->
**Leaders are managers**. There is a history within managerial studies among WEIRD populations of contrasting *leaders* from *managers*. @zaleznik_managers_1977, for example, suggested managers seek systematicity, organization, and stability, whereas leaders are willing to accept a lack of control and creatively navigate uncertainty to produce novel insight and appeale to the emotions of followers. Political anthropology reveals, however, in the absence of state or organizational institutions, local leaders are often tasked with both managing various economic systems while also providing social services and interacting with followers on an inter-personal level [@cohen_comparative_1967]. Within the domain of leader management, there are likely to be two primary features of human leader functions: strategizing and organizing. Our results correspond very closely to what @van_vugt_cognitive_2007 define as *strategic leadership*, where leaders increase the pay-offs of another individuals behavior thereby incentivizing followership versus *coordinating leadership*, where individuals bound by group challenges must effectively organize collective effort.

@@

<!-- PROSCIAL FUNCTIONS FEATURE -->
**Public goods**. Beyond strategically shaping inter-individual behaviors or organizing collective behavior, we found much of leadership involves beneficial investments in group welfare. The prosocial feature of leader functions included distributing resources, providing protection, and various prosocial investments, which are functions commonly highlighted in the anthropological literature on leadership in nonindustrial societies [@sahlins_poor_1963; @johnson_organizational_1982; @service_origins_1975; @boehm_egalitarian_1993; @boehm_hierarchy_1999; @johnson_evolution_1987]. In the absence of institutionalized leadership roles, which characterize many of the relatively small-scale populations in this sample, leaders often maintain social influence via continued community investment and the demands of followers can shape the functions of leaders to meet local needs.

@@

# Concluding remarks: Beyond the non-WEIRD 
<!-- NOT SURE OF TITLE -- WANT TO EMPHASIZE NON WEIRD DIVERSITY AND ALSO MENTION BACK TO WERID FINDINGS BRIEFLY -->
In recognizing the limitations of most extant psychological, sociological, and managerial data, many social scientists are now diversifying their samples to better understand *human* psychology, behavior, and culture. We caution that the WEIRD people problem in behavioral science cannot be remedied simply by a "non-WEIRD" solution. As our results, as well as much of anthropology and ethnography, demonstrate, within the populations that are now often categorized as non-WEIRD, there is enormous diversity. Furthermore, larger order categories such as "subsistence strategy" are probably less relevant for understanding behavioral diversity than more salient group contexts. Despite limitations, the ethnographic record presents an invaluable opportunity to explore patterns of human diversity and evolutionary theorists and researchers looking beyond WEIRD samples should take special interest in comparative ethnographic methods.

We found leaders in the ethnographic record are portrayed as generally prosocial, serve to resolve conflicts and organize cooperation, and receive material, social, and reproductive benefits. These results conform to the characteristics of leaders that are emphasized in the literature on the evolution of leadership and provide broad cross-cultural support for several influential theories, such as the prestige-dominance model, collective action models, the service-for-prestige model, and the computational services model [for review, see @garfield_evolutionary_2019].

While there is a evidence of a dominace style of leadership involving coercive authority, confidence and aggression, as well as a prestige style involving special knowledge, experience, and innovation, we suggest rethinking dichotomous models of leadership styles and focusing on inter-individual variation in specific leadership competencies. There is likely to be substantial overlap in physical, spiritual, cognitive, and other prosocial endowments and the degree to which a leader relies on any will be shaped by their individual levels of such capacities, the presence and capacities of their rivals, as well as group context, their functions, and cultural norms. For example, widespread preferences for egalitarianism and social equality are likely strong drivers of followers preferences for leaders across diverse societies [@cheng_dominance_2019], but shamanism and ritual based contexts represent a social domain in which ideologies support a style of leadership fusing both dominance and prestige-based competencies. We suspect cultural and pseudoreligous ideologies also play important roles in shaping leadership styles follower preferences among WEIRD populations and in national political movements around the world.

>>>>

Overall, our results suggest that although high status leaders are common across cultures and all levels of social organization, only some leaders are high status, namely, those with a limited set of qualities such as wealth, experience, decision-making abilities, intelligence, and interpersonal skills, as well as older age and conformity to social norms.

@@

**Dominance and prestige**. Much of the evolutionary theoretical literature on leadership has suggested social influence is most often achieved via one of two distinct strategies: dominance or prestige [for review see @garfield_review_2019; @garfield_evolutionary_2019]. There is a long history of leadership styles being described as based on either force and strength (i.e., dominance) or respect and special skills (i.e., prestige). Such dual-models are rooted in the coupling of ethology and animal dominance hierarchies with ethnography and cultural prestige systems [see @mead_sex_1935; @sahlins_poor_1963; @lewis_leaders_1974; @tiger_dominance_1970; @bernstein_dominance_1976]. Synthesizing the ethnographic and ethological evidence, @kracke_force_1978 linked "force and persuassion" to psychological-affect between leaders and followers in small-scale societies and @barkow_prestige_1980 suggested the evolution of human prestige was linked to sexual selection and symbolic culture. The dominance-prestige theory as formulated by @henrich_evolution_2001 significantly advanced evolutionary approaches to leadership and social hierarchy and inspired much subsequent work across the behavioral sciences [e.g., @van_vugt_dual_2019; @cheng_two_2013; @maner_2016; @halevy_status_2012; @suessenbach_dominance_2019; @maner_dominance_2017; @garfield_evolutionary_2019; @garfield_investigating_2019]. @van_vugt_dual_2019 [p. 1], for example, "argue that there are two leadership styles: prestige style and dominance style," and @cheng_two_2013 [p. 119] in focusing on distinctions between dominance and prestige suggests, "human social stratification is characterized by the co-occurrence of both strategies." Exploratory cluster analyses reported here, however, suggest multiple features of leader qualities, beyond dominance and prestige.

<!-- DISCUSSES COMPETENCIES FEATURE -->
**Leader capacities**. The individual-level competencies feature was comprised of variables associated with dominance-based leadership, including coercive authority and aggressiveness [@tiger_imperial_1971; @barkow_prestige_1975; @barkow_darwin_1989; @henrich_evolution_2001; @cheng_two_2013] as well as the qualities of knowledgeable or intelligent and experienced or accomplished, traits that have been emphasized by the prestige theory [@henrich_evolution_2001; @cheng_two_2013] and the more general computational services theory [@hagen_leadership_2019; @garfield_evolutionary_2019], and which are among the most strongly supported in the ethnographic record (Figure \@ref(fig:varsupport)). Interestingly, being feared and having supernatural capacities clustered more closely with the prestige-style traits than with the dominance-style traits. There is likely a strong connection between shamanism and ritual-based influence with both prestige and dominance styles of leadership. Ritual and supernatural capacities may represent a dimension of leader qualities in which dominance and prestige are coupled forming an alternative style of leadership, as shamans often provide valued public good, have special expertise and abilities, are feared, and achieve or maintain influence via threats or use of direct and/or indirect aggression. Artistic performance is also implicated in this style of leadership and ceremonial, artistic performances by shamans are common displays which often incorporate the superhuman abilities of shamans [@singh_cultural_2017].

As brief examples of the diversity of leader capacities, of the very many in the ethnographic record, among the Azande, "though the most knowledgeable witch-doctors tend to become leaders, there is no institutionalized organization or leadership among them" [@fo07-056, p. 85]; and among the Highland Scots, @es10-003 [p. 147] notes, "on the whole, their long ties with place and people, coupled with education, made them useful leaders of a community in which leaders were needed to adapt life to a changing economy." Such exceptionally talented leaders, however, also face unique challenges, particularly within more egalitarian societies which emphasizes personal autonomy and resit hierarchy. @nm09-056 [p. 183] describes Laura Cornelius Kellogg (also known as Minnie and Wynnogene, 1880-1947), a land-claims activist, writer, and perhaps the most famous female leader among the Iroquois:

@@

<!-- DISCUSSES PROSOCIALITY FEATURE -->
**Prosocial norms**. The prosociality feature of leader qualities, including generosity, interpersonal and oratory skills, and charisma as well as the role of cultural norms in shaping leadership styles, including proper behavior and having a favorable personality, generally correspond to a charismatic style of leadership that has been identified in industrialized settings and widely discussed in sociology and more recently emphasized in evolutionary psychology models of leadership [@den_hartog_culture_1999; @weber_distribution_1978; @grabo_charismatic_2016; @grabo_charisma_2017; @howell_role_2005]. The importance of prosociality  also corresponds to the @henrich_big_2015 Big-man mechanism, which suggests positive assortment of followers to prosocial leaders can facilitate the evolution of cooperative social norms and prestige-based leadership.
<!-- A few variables did not seem to fit prosocial theme, however, such as *Physical health*, *Physical formidable*, and *Drug consumption* (although drug use is often an important social activity). -->

@@

<!-- DISCUSSES BOTH MANAGEMENT FEATURES -->
**Leaders are managers**. There is a history within managerial studies among WEIRD populations of contrasting *leaders* from *managers*. @zaleznik_managers_1977, for example, suggested "managers" seek systematicity, organization, and stability, whereas "leaders" are willing to accept a lack of control and creatively navigate uncertainty to produce novel insight and appeal to the emotions of followers. Political anthropology reveals, however, in the absence of state-based or large-scale institutions, local leaders are often tasked with both managing various economic systems while also providing social services and interacting with followers on an interpersonal level [@cohen_comparative_1967]. Within the domain of leader management, there are likely to be two primary features of human leader functions: strategizing and organizing. Our results correspond very closely to what @van_vugt_cognitive_2007 define as *strategic leadership*, where leaders increase the pay-offs of another individuals behavior thereby incentivizing followership versus *coordinating leadership*, where individuals bound by group challenges must effectively organize collective effort.

@@

<!-- PROSCIAL FUNCTIONS FEATURE -->
**Public goods**. Beyond strategically shaping inter-individual behaviors or organizing collective behavior, we found much of leadership involves beneficial investments in group welfare. The prosocial feature of leader functions included distributing resources, providing protection, and various prosocial investments, which are functions commonly highlighted in the anthropological literature on leadership in nonindustrial societies [@sahlins_poor_1963; @johnson_organizational_1982; @service_origins_1975; @boehm_egalitarian_1993; @boehm_hierarchy_1999; @johnson_evolution_1987]. In the absence of institutionalized leadership roles, which characterize many of the relatively small-scale populations in this sample, leaders often maintain their social influence via continued community investment and the demands of followers can shape the functions of leaders to meet local needs.

@@

# Concluding remarks: Beyond the non-WEIRD 
<!-- NOT SURE OF TITLE -- WANT TO EMPHASIZE NON WEIRD DIVERSITY AND ALSO MENTION BACK TO WERID FINDINGS BRIEFLY -->
In recognizing the limitations of most extant psychological, sociological, and managerial data, many social scientists are now diversifying their samples to better understand *human* psychology, behavior, and culture. We caution that the WEIRD people problem in behavioral science cannot be remedied simply by a "non-WEIRD" solution. Our results, and much of anthropology and ethnography, clearly demonstrate enormous diveristy within populations that are now often categorized as non-WEIRD. Furthermore, higher-order categories such as "subsistence strategy" are probably less relevant for understanding the complexity of behavioral diversity than more salient group contexts. Despite limitations, the ethnographic record presents an invaluable opportunity to explore patterns of human cultural and behavioral diversity. Evolutionary theorists and researchers looking beyond WEIRD samples should take special interest in comparative ethnographic methods.

We found leaders in the ethnographic record are portrayed as generally prosocial, are equipped with special capacities, serve to resolve conflicts and organize cooperation, and receive material, social, and reproductive benefits. These results generally provide broad cross-cultural support for several influential theories, such as the prestige-dominance model, collective action models, the service-for-prestige model, and the computational services model [for review, see @garfield_evolutionary_2019].

While there is evidence of a dominace style of leadership involving coercive authority, confidence and aggression, as well as a prestige style involving special knowledge, experience, and innovation, we suggest rethinking dichotomous models of leadership styles and focusing on inter-individual variation in specific leadership competencies. There is likely to be substantial overlap in physical, spiritual, cognitive, and other prosocial endowments and the degree to which a leader relies on any of these capacities will be shaped by their individual levels of such capacities, the capacities of ptential rivals, as well as group context, their functions, and cultural norms. For example, widespread preferences for egalitarianism and social equality are likely strong drivers of followers preferences for leaders across diverse societies [e.g., @cheng_dominance_2019; @price_evolution_2014], but shamanism and ritual based contexts represent a social domain in which ideologies support a style of leadership fusing both dominance and prestige-based competencies. It seems evident that cultural forces and ideologies play important roles in shaping leadership styles and follower preferencesm including among WEIRD populations and in national political movements around the world [e.g., @mcdermott_leadership_2018; @inglehart_trump_2016].
